Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-10-18 anderson-etal-blackswan-timeseries.tex
Clean up main .tex file
Wrap ms and (very) minor edits

\documentclass[12pt]{article}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2.54cm,bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
\geometry{letterpaper}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{lineno}
\usepackage{ragged2e}
\usepackage{color}
\usepackage{verbatim}
\usepackage[sectionbib]{chapterbib}

\usepackage{amsmath}
% \makeatletter
% \g@addto@macro\normalsize{%
%   \setlength\abovedisplayskip{-20pt}
%   \setlength\belowdisplayskip{0pt}
%   \setlength\abovedisplayshortskip{-20pt}
%   \setlength\belowdisplayshortskip{0pt}
% }
% \makeatother

\usepackage{longtable}
\usepackage{multirow}
\usepackage{array}
\usepackage{tabu} % for spacing between rows in longtable
\setlength{\tabulinesep}{9pt}

@@

\title{Black swans in ecological time series\\Black swans in population dynamics\\Heavy-tailed ecological time series\\Black swans in ecological dynamics}
\author{
Sean C. Anderson$^{1\ast}$ \and
Trevor A. Branch$^2$ \and
Andrew B. Cooper$^3$ \and
Nicholas K. Dulvy$^1$
}
\date{}

@@

\setlength\parskip{0.16in}
\setlength\parindent{0in}
\usepackage{titlesec}
\titlespacing\section{0pt}{6pt plus 4pt minus 2pt}{-6pt plus 2pt minus 2pt}
\titlespacing\subsection{0pt}{0pt plus 4pt minus 2pt}{-8pt plus 2pt minus 2pt}

@@

\maketitle

@@

\emph{Authors are listed alphabetically (ABCD!) and open to reordering.}

\textsuperscript{1}Earth to ocean research group, department of biological sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada

\textsuperscript{2}School of Aquatic and Fishery Sciences, University of Washington, Box 355020, Seattle, WA 98195, USA

\textsuperscript{3}School of Resource and Environmental Management, Simon Fraser University, Burnaby, BC, V5A 1S6, Canada

\textsuperscript{*}Corresponding author: Sean C. Anderson; Earth to Ocean Research Group, Department of Biological Sciences, Simon Fraser University, Burnaby BC, V5A 1S6; Phone: 1-778-782-3989; E-mail: sean\_anderson@sfu.ca

%\begin{spacing}{1.4}
\doublespacing
%\onehalfspacing

>>>>

\documentclass[12pt]{article}
\usepackage{geometry}
\geometry{verbose, letterpaper, tmargin = 2.54cm, bmargin = 2.54cm,
  lmargin = 2.54cm, rmargin = 2.54cm}
\geometry{letterpaper}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{lineno}
\usepackage{ragged2e}
\usepackage{color}
\usepackage{verbatim}
\usepackage[sectionbib]{chapterbib}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage{multirow}
\usepackage{array}
\usepackage{tabu} % for spacing between rows in longtable
\setlength{\tabulinesep}{9pt}

@@

\title{Black swans in ecological time series\\
  Black swans in population dynamics\\
  Heavy-tailed ecological time series\\
  Black swans in ecological dynamics\\
  Black swan ecological dynamics\\
  \ldots}
\author{
Sean C. Anderson$^{1\ast}$ \and
Trevor A. Branch$^2$ \and
Andrew B. Cooper$^3$ \and
Nicholas K. Dulvy$^1$
}
\date{}

@@

@@

\maketitle
\doublespacing
%\onehalfspacing

@@

\noindent
\emph{Authors are listed alphabetically (ABCD!) and open to reordering.}

\noindent
\textsuperscript{1}Earth to ocean research group,
Department of biological sciences,
Simon Fraser University, Burnaby BC, V5A 1S6, Canada

\noindent
\textsuperscript{2}School of Aquatic and Fishery Sciences,
University of Washington, Box 355020, Seattle, WA 98195, USA

\noindent
\textsuperscript{3}School of Resource and Environmental Management,
Simon Fraser University, Burnaby, BC, V5A 1S6, Canada

\noindent
\textsuperscript{*}Corresponding author: Sean C. Anderson;
Earth to Ocean Research Group,
Department of Biological Sciences,
Simon Fraser University,
Burnaby BC, V5A 1S6;
Phone: 1-778-782-3989;
E-mail: sean\_anderson@sfu.ca

### 2014-10-18 ms.tex
Wrap ms and (very) minor edits

Black swans are statistically improbable events that nonetheless occur --- often with profound implications. While extremes in the physical environment, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency, it remains unclear the extent to which ecological populations buffer or suffer from extremes. Here, we develop a probability model to estimate the degree of heavy-tailedness (presence of black swans) in ecological process noise. We apply our model to \NPops~time series from around the world across \NOrders~taxonomic orders and seven classes. We find strong evidence of black swans, but they are rare, occurring in \overallMinPerc--\overallMaxPerc\% of populations; most frequently for birds (\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects (\InsectaRangePerc\%), and fishes (\OsteichthyesRangePerc\%). When they occur, they tend to be driven by climate and severe winters, cycles of parasites and predators, and interactions between these elements. They are more frequently detected for populations with longer time series and lower levels of process noise. We find little evidence of intrinsic life-history covariates. \emph{close with modelling and policy implications}

@@

There are two key reasons why we may find little evidence of ecological black swans. First, they might not exist. Indeed, the vast majority of population dynamics model fitting and risk forecasting assumes that population dynamics are not heavy tailed (e.g.~REF, REF, REF). Alternatively, population dynamics might be heavy tailed, but a paucity of statistical methods and shortness of available time series has limited our ability to detect them. For example, a commonly applied probability distribution to detect the shape of time series tails (the generalized extreme value distribution) requires lumping data into the most extreme event per unit time (e.g.~heaviest rainfall per year) and therefore requires a considerable quantity of data (REF). Previous work into the shape of population dynamic tails, has focused on model comparison to identify parsimonious explanations from a set of particular hypotheses, but does not quantify the magnitude of heavy tails on a continuous scale or put the conclusions in a probabilistic framework (REF). Furthermore, there is evidence that information criteria are poor at distinguishing subtle differences in population dynamics for the relatively short time series that are most readily available (REF).

@@

There a number of caveats when considering the generality of our results. First, the GPDD data represent a taxonomically and geographically biased sample of populations. Also, the longer time series we focus on are dominated by commercially and recreationally important species and a disproportionate number of populations are located in the United Kingdom. Although we expect that we would find strong but rare evidence for black swans in other large taxonomic or geographic samples, the common forces driving those black swans (e.g.~severe winters in Table~1) likely differ. Second, in a large database with disparate datasources, we cannot ensure the quality of every datapoint. Recording mistakes undoubtedly comprise some of the black swans we detected. Conversely, recording mistakes could mask heavy tails if initial observations are discarded or altered as assumed recording errors. Third, the temporal scale of observation and population dynamics vary considerably across populations in the GPDD and this likely influences the detection of heavy tails. For example, if we make frequent observations relative to generation time (e.g.~for many large bodied mammals) we will average across generations and perhaps not observe the tails. Furthermore, if we record population censuses infrequently relative to generation time (e.g.~many insects in the GPDD) the recorded data may average across extreme and less-extreme events and dampen observed black swan dynamics.

>>>>

Black swans are statistically improbable events that nonetheless occur --- often
with profound implications. While extremes in the physical environment,such as
monsoons and heat waves, are widely studied and increasing in magnitude and
frequency, it remains unclear the extent to which ecological populations buffer
or suffer from extremes. Here, we develop a probability model to estimate the
degree of heavy-tailedness (presence of black swans) in ecological process
noise. We apply our model to \NPops~time series from around the world across
\NOrders~taxonomic orders and seven classes. We find strong evidence of black
swans, but they are rare, occurring in \overallMinPerc--\overallMaxPerc\%of
populations; most frequently for birds (\AvesRangePerc\%) followed by mammals
(\MammaliaRangePerc\%), insects (\InsectaRangePerc\%), and
fishes(\OsteichthyesRangePerc\%). When they occur, they tend to be driven by
climate and severe winters, cycles of parasites and predators, and interactions
between these elements. They are more frequently detected for populations with
longer time series and lower levels of process noise. We find little evidence
of intrinsic life-history covariates. TODO close with modelling and policy
implications

@@

There are two key reasons why we may find little evidence of ecological black
swans. First, they might not exist. Indeed, the majority of population
dynamics model fitting and risk forecasting assumes that population dynamics
are not heavy tailed (e.g.~REF, REF, REF). Alternatively, population dynamics
might be heavy tailed, but a paucity of statistical methods and shortness of
available time series has limited our ability to detect them. For example,
a commonly applied probability distribution to detect the shape of time series
tails (the generalized extreme value distribution) requires lumping data into
the most extreme event per unit time (e.g.~heaviest rainfall per year) and
therefore requires a considerable quantity of data (REF). Previous work into
the shape of population dynamic tails, has focused on model comparison to
identify parsimonious explanations from a set of particular hypotheses, but
does not quantify the magnitude of heavy tails on a continuous scale or put the
conclusions in a probabilistic framework (REF). Furthermore, information
criteria are poor at distinguishing subtle differences in population dynamics
for the relatively short time series that are most readily available (REF CHANGE
THIS?).

@@

There a number of caveats when considering the generality of our results.
First, the GPDD data represent a taxonomically and geographically biased sample
of populations. Also, the longer time series we focus on are dominated by
commercially and recreationally important species and a disproportionate number
of populations are located in the United Kingdom. Although we expect that we
would find strong but rare evidence for black swans in other large taxonomic or
geographic samples, the common forces driving those black swans (e.g.~severe
winters in Table~1) likely differ. Second, in a large database with disparate
data sources, we cannot ensure the quality of every data point. Recording
mistakes undoubtedly comprise some of the black swans we detected. Conversely,
recording mistakes could mask heavy tails if initial observations are discarded
or altered as assumed recording errors. Third, the temporal scale of observation
and population dynamics vary considerably across populations in the GPDD and
this likely influences the detection of heavy tails. For example, if we make
frequent observations relative to generation time (e.g.~for many large bodied
mammals) we will average across generations and perhaps not observe the tails.
Furthermore, if we record population censuses infrequently relative to
generation time (e.g.~many insects in the GPDD) the recorded data may average
across extreme and less-extreme events and dampen observed black swan dynamics.

### 2014-10-18 supp.tex
Wrap ms and (very) minor edits

\begin{enumerate}
\item
  To remove populations with unreliable population indices that could be strongly confounded with economics and sampling effort, we removed all populations with a sampling protocol listed as \texttt{harvest} as well populations with the words \texttt{harvest} or \texttt{fur} in the cited reference title.
\item
  We removed all populations with uneven sampling intervals. I.e.~we removed populations that didn't have a constant difference between the ``decimal year begin'' and ``decimal year end'' columns.
\item
  We removed all populations rated as $< 2$ in the GPDD quality assessment (on a scale of $1$ to $5$, with $1$ being the lowest quality data) \citep[following][]{sibly2005, ziebarth2010}
\item
  Populations with negative abundance values were assumed to be log values and transformed by taking $10$ to the power of the recorded abundance. In many cases this was noted for the population, but not in all cases. We inspected each of these \totalAssumedLog~time series to make sure our assumption made sense.
\item
  We filled in all missing time steps with \texttt{NA} values and imputed single missing values with the geometric mean of the previous and following values. We chose a geometric mean to be linear on the log scale that the Gompertz and Ricker-logistic models were fitted on.
\item
  We filled in single recorded values of zero with the lowest non-zero value in the time series \citep[following][]{brook2006a}. This assumes that single values of zero result from abundance being low enough that censusing missed present individuals. We turned multiple zero values in a row into \texttt{NA} values. This implies that multiple zero values were either censusing errors or caused by emigration. Regardless, our population models were fitted on a multiplicative (log) scale and so could not account for zero abundance.
\item
  We removed all populations with four or more identical values in a row since these suggest either recording error or extrapolation between two observations.
\item
  We removed all populations without at least four unique values \citep[following][]{brook2006a}.
\item
  We then wrote an algorithm to find the longest unbroken window of abundance (no \texttt{NA}s) with at least $20$ time steps in each population time series. If there were any populations with multiple windows of identical length, we took the most recent window. This is a longer window than used in some previous analyses \citep[e.g.][]{brook2006a}, but since our model attempts to capture the shape of the distribution tails, our model requires more data.
\item
  Finally, we removed GPDD Main ID \texttt{20531}, which we noticed was a duplicate of \texttt{10139} (a heron population).
\end{enumerate}
\noindent
We provide a supplemental figure of all the time series included in our analysis and indicate which values were interpolated (\percImputedPops\% of populations had at least one point interpolated and only \percImputedPoints\% of the total observations were interpolated) (Fig.~\ref{fig:all-ts}). Table S1 shows the final taxonomic breakdown and the number of populations with interpolated values.

@@

Our prior on $\nu$ is based on \citet{fernandez1998}; they chose a more informative exponential rate parameter of $0.1$. We chose a less informative rate parameter of $0.01$ and truncated the distribution at $2$, since at $\nu < 2$ the variance of the t distribution is undefined. This prior gives only a $7.7$\% probability $\nu < 10$ but constrains the sampling sufficiently to avoid wandering off towards infinity --- above approximately $\nu = 20$ the t distribution is so similar to the normal distribution (Fig.~\ref{fig:didactic}) that time series of the length considered here are unlikely to be sufficiently informative about the precise value of $\nu$. In the scenario where the data are not heavy tailed (e.g.~Fig.~\ref{fig:didactic}e, h) the posterior will approximately match the prior (median $= 71$, mean $= 102$) and not be flagged as likely heavy tailed using the metrics we used in our paper (e.g.~Pr$(\nu < 10) > 0.5$).

@@

We incorporated weakly informative priors into our model: $\mathrm{Cauchy}(0, 10)$ on the global intercept $\alpha$, $\mathrm{Half\mhyphen Cauchy}(0, 2.5)$ on all standard deviation parameters, $\mathrm{Half\mhyphen Cauchy}(0, 10)$ on the dispersion parameter $\phi_\mathrm{disp}$, and $\mathrm{Cauchy}(0, 2.5)$ on all other parameters \citep{gelman2006c,gelman2008d}. Compared to normal priors, the Cauchy priors concentrate more probability density at reasonably parameter values while allowing for higher probability density far into the tails, thereby allowing the data to dominate the posterior more strongly if it disagrees with the prior. Our conclusions were not qualitatively changed by using uniform priors. We fitted our models with 5000 total iterations per chain, 2500 warm-up iterations, four chains, and discarding every second sample to save memory. We checked for chain convergence visually and with the same criteria as before ($\widehat{R} < 1.05$ and $n_\mathrm{eft} >200$ for all parameters).

>>>>

\begin{enumerate}

\item To remove populations with unreliable population indices that could be
  strongly confounded with economics and sampling effort, we removed all
  populations with a sampling protocol listed as \texttt{harvest} as well
  populations with the words \texttt{harvest} or \texttt{fur} in the cited
  reference title.

\item We removed all populations with uneven sampling intervals. I.e.~we removed
  populations that didn't have a constant difference between the ``decimal year
  begin'' and ``decimal year end'' columns.

\item We removed all populations rated as $< 2$ in the GPDD quality assessment
  (on a scale of $1$ to $5$, with $1$ being the lowest quality data)
  \citep[following][]{sibly2005, ziebarth2010}

\item Populations with negative abundance values were assumed to be log values
  and transformed by taking $10$ to the power of the recorded abundance. In many
  cases this was noted for the population, but not in all cases. We inspected
  each of these \totalAssumedLog~time series to make sure our assumption made
  sense.

\item We filled in all missing time steps with \texttt{NA} values and imputed
  single missing values with the geometric mean of the previous and following
  values. We chose a geometric mean to be linear on the log scale that the
  Gompertz and Ricker-logistic models were fitted on.

\item We filled in single recorded values of zero with the lowest non-zero value
  in the time series \citep[following][]{brook2006a}. This assumes that single
  values of zero result from abundance being low enough that censusing missed
  present individuals. We turned multiple zero values in a row into \texttt{NA}
  values. This implies that multiple zero values were either censusing errors or
  caused by emigration. Regardless, our population models were fitted on
  a multiplicative (log) scale and so could not account for zero abundance.

\item We removed all populations with four or more identical values in a row
  since these suggest either recording error or extrapolation between two
  observations.

\item We removed all populations without at least four unique values
  \citep[following][]{brook2006a}.

\item We then wrote an algorithm to find the longest unbroken window of
  abundance (no \texttt{NA}s) with at least $20$ time steps in each population
  time series. If there were any populations with multiple windows of identical
  length, we took the most recent window. This is a longer window than used in
  some previous analyses \citep[e.g.][]{brook2006a}, but since our model
  attempts to capture the shape of the distribution tails, our model requires
  more data.

\item Finally, we removed GPDD Main ID \texttt{20531}, which we noticed was
  a duplicate of \texttt{10139} (a heron population).

\end{enumerate}

\noindent
We provide a supplemental figure of all the time series included in our analysis
and indicate which values were interpolated (\percImputedPops\% of populations
had at least one point interpolated and only \percImputedPoints\% of the total
observations were interpolated) (Fig.~\ref{fig:all-ts}). Table S1 shows the
final taxonomic breakdown and the number of populations with interpolated
values.

@@

Our prior on $\nu$ was based on \citet{fernandez1998}; they chose a more
informative exponential rate parameter of $0.1$. We chose a less informative
rate parameter of $0.01$ and truncated the distribution at $2$, since at $\nu
< 2$ the variance of the t distribution is undefined. This prior gives only
a $7.7$\% probability $\nu < 10$ but constrains the sampling sufficiently to
avoid wandering off towards infinity --- above approximately $\nu = 20$ the
t distribution is so similar to the normal distribution
(Fig.~\ref{fig:didactic}) that time series of the length considered here are
unlikely to be sufficiently informative about the precise value of $\nu$. In the
scenario where the data are not heavy tailed (e.g.~Fig.~\ref{fig:didactic}e, h)
the posterior will approximately match the prior (median $= 71$, mean $= 102$)
and not be flagged as likely heavy tailed using the metrics we used in our paper
(e.g.~Pr$(\nu < 10) > 0.5$).

@@

We incorporated weakly informative priors into our model: $\mathrm{Cauchy}(0,
10)$ on the global intercept $\alpha$, $\mathrm{Half\mhyphen Cauchy}(0, 2.5)$ on
all standard deviation parameters, $\mathrm{Half\mhyphen Cauchy}(0, 10)$ on the
dispersion parameter $\phi_\mathrm{disp}$, and $\mathrm{Cauchy}(0, 2.5)$ on all
other parameters \citep{gelman2006c, gelman2008d}. Compared to normal priors,
the Cauchy priors concentrate more probability density at reasonably parameter
values while allowing for higher probability density far into the tails, thereby
allowing the data to dominate the posterior more strongly if it disagrees with
the prior. Our conclusions were not qualitatively changed by using uniform
priors. We fitted our models with 5000 total iterations per chain, 2500 warm-up
iterations, four chains, and discarding every second sample to save memory. We
checked for chain convergence visually and with the same criteria as before
($\widehat{R} < 1.05$ and $n_\mathrm{eft} >200$ for all parameters).
